---
title: "Introduction to Stochastic Compartmental Models"
subtitle: "STAT 244NF: Infectious Disease Modeling"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)

lambda_t_fcn <- function(R_0, D, I_i, N){
  c_e <- R_0/D
  return(1-exp(-c_e*I_i/N))
}

SEIR_simulation <- function(N, E0, I0, R0, R_0, pD, D, Time){
  SEIR_df <- data.frame(time=0:Time,
                     S=rep(NA, Time+1),
                     E=rep(NA, Time+1),
                     I=rep(NA, Time+1),
                     R=rep(NA, Time+1),
                     lambda_t=rep(NA, Time+1))

  SEIR_df$S[1] <- N-E0-I0-R0
  SEIR_df$E[1] <- E0
  SEIR_df$I[1] <- I0
  SEIR_df$R[1] <- R0

  for (i in 2:(Time+1)){
    SEIR_df$lambda_t[i] <- lambda_t_fcn(R_0=R_0, D=D, I_i=SEIR_df$I[i-1], N=N)
    SEIR_df$S[i] <- SEIR_df$S[i-1]-SEIR_df$lambda_t[i]*SEIR_df$S[i-1]
    SEIR_df$E[i] <- SEIR_df$E[i-1]+SEIR_df$lambda_t[i]*SEIR_df$S[i-1]-1/pD*SEIR_df$E[i-1]
    SEIR_df$I[i] <- SEIR_df$I[i-1]+1/pD*SEIR_df$E[i-1]-1/D*SEIR_df$I[i-1]
    SEIR_df$R[i] <- SEIR_df$R[i-1]+1/D*SEIR_df$I[i-1]
  }
  
  return(list(
    df_ggplot=data.frame(time=rep(0:Time, 4),
                       compartment=rep(c("S","E", "I", "R"), each=(Time+1)),
                       count=c(SEIR_df$S, SEIR_df$E, SEIR_df$I, SEIR_df$R)),
    SEIR_df=SEIR_df)
  )
}
```

# Deterministic models

So far in this class, we have been working with deterministic models. Recall from our introduction to compartmental models, deterministic models describe what happens "on average" in a population. Parameters (e.g., $\lambda_t$, $\pi$, $\rho$) are fixed, and model predictions are predetermined. In other words, there is no randomness. Even though $\lambda_t$ changes based on the value of $I_{t-1}$, it will always be the same for each value of $t$, no matter how many times we run our simulation. Similarly, $\pi$ and $\rho$ will always be the same.  

For example, consider the simulation from the SEIR model: 

```{r}
sim_SEIR <- SEIR_simulation(N=100, E0=0, I0=1, R0=0, R_0=2, pD=3, D=2, Time=60)

head(sim_SEIR$SEIR_df, 10)
```

The values above represent our predictions at the first nine days (plus the initial day), assuming a population of 100 individuals, 1 initial infectious individuals, and an infection with a basic reproductive number of 2, a pre-infectious period of 3 days, and an infectious period of 2 days. 

**Do you think it is reasonable to predict that 1.25 individuals will be infectious at Day 7? If not, what would be more reasonable?**

\vspace{1.5cm}

**Do you think it is reasonable that the size of the outbreak will always be the same? If not, what would be more reasonable?**

\vspace{1.5cm}

## Difference Equations 

\begin{align*}
S_t&=S_{t-1}-\underbrace{{\color{green}{\lambda_t S_{t-1}}}}_{\text{not an integer}}\\
E_t&=E_{t-1}+{\color{green}{\lambda_tS_{t-1}}}-\underbrace{{\color{blue}{\pi E_{t-1}}}}_{\text{not an integer}}\\
I_t&=I_{t-1}+{\color{blue}{\pi E_{t-1}}}-\underbrace{{\color{violet}{\rho I_{t-1}}}}_{\text{not an integer}}\\
R_t&=R_{t-1}+{\color{violet}{\rho I_{t-1}}}\\
\end{align*}

```{r fig.align='center'}
ggplot(data=sim_SEIR$df_ggplot, aes(x=time, y=count)) + 
  geom_line(aes(color=compartment), size=1.1) + 
  theme_bw() + 
  xlab("Days") + 
  ylab("Count") + 
  scale_colour_discrete(limits = c("S", "E", "I", "R"))
```

## Problems with the deterministic models

1. Reasonable predictions for compartment counts should have only integer-valued - to be realistic, we have to have whole counts.
2. Random chance plays a role in outbreak size. We can conceivably have the same population and in one case we would get an outbreak, but in another we would not.

**These problems can be remedied through the use of stochastic models.**


# Stochastic models

In stochastic models, two things are different than the deterministic models we were using. First, we assume that the number of individuals moving between compartments is a random variable that follows a particular probability distribution. Second, the number of individuals that move between compartments varies by chance. A range of outcomes are possible, which is more representative of the real world. These two qualities address the two problems listed above. 

## Difference Equations 

\begin{align*}
S_t&=S_{t-1}-{\color{green}{E^{(SE)}_{t}}}\\
E_t&=E_{t-1}+{\color{green}{E^{(SE)}_{t}}}-{\color{blue}{I^{(EI)}_t}}\\
I_t&=I_{t-1}+{\color{blue}{I^{(EI)}_t}}-{\color{violet}{R^{(IR)}_t}}\\
R_t&=R_{t-1}+{\color{violet}{R^{(IR)}_t}}\\
\end{align*}

## Random Variables

Recall from your introductory statistics class, a *random variable* is a random process with a numerical outcome; usually represented by a capital letter like X, Y, Z. 

Here we are going to be more specific in our notation for each random variable:

- ${\color{green}{E^{(SE)}_{t}}}$: number of newly pre-infectious individuals at time $t$
- ${\color{blue}{I^{(EI)}_t}}$: number of newly infectious individuals at time $t$
- ${\color{violet}{R^{(IR)}_t}}$: number of newly recovered individuals at time $t$

**Each random variable follows some probability distribution. What are some probability distributions that we know?**

\vspace{1.5cm}

**Which of these distributions will have integer outcomes?**

\vspace{1.5cm}

\newpage

- ${\color{green}{E^{(SE)}_{t}}} \sim$
- ${\color{blue}{I^{(EI)}_t}} \sim$
- ${\color{violet}{R^{(IR)}_t}} \sim$

\newpage

## Example 

Consider an SEIR model with the following:

- $N=50$ (population is 50)
- $S(0)=49$ (49 people are susceptible to start)
- $E(0)=0$
- $I(0)=1$ (1 person is infectious to start)
- $R(0)=0$
- $R_0=2$ (the basic reproductive number is 2)
- $D'=2$ days (the pre-infectious period is 2 days); $\pi=1/D'$
- $D=2$ days (the infectious period is 2 days); $\rho=1/D$
- Time=15 days (total number of days we are observing this for purposes of this example)

For the values listed above, follow these steps for $t=1,...,15$:

1. Calculate the risk of infection in the next time interval using $\lambda_t=1-\exp\left\{-c_e\frac{I_{t-1}}{N} \right\}$.
2. Randomly sample $E^{(SE)}_t \sim Binomial\left(S_{t-1}, \lambda_t \right)$. Update your value for $S_t$. You can use `rbinom(n=1, size=S[t-1], prob=lambda_t)`. Add your point to the plot below.
3. Randomly sample $I^{(EI)}_t \sim Binomial\left(E_{t-1}, \pi \right)$. Update your value for $E_t$ (you can use `rbinom(n=1, size=E[t-1], prob=pi)`. Add your point to the plot below.
4. Randomly sample $R^{(IR)}_t \sim Binomial(I_{t-1}, \rho)$. Update your values for $I_t$ and $R_t$, and add the points to your plot below.

Once you are done, add your points to the plots on the blackboard.

```{r fig.align='center', echo=FALSE}
sim_SEIR <- SEIR_simulation(N=50, E0=0, I0=1, R0=0, R_0=3, pD=2, D=2, Time=15)

ggplot(data=sim_SEIR$df_ggplot, aes(x=time, y=count)) + 
  # geom_line(aes(color=compartment), size=1.1) +
  theme_bw() + 
  xlab("Days") + 
  ylab("Count") + 
  scale_colour_discrete(limits = c("S", "E", "I", "R"))
```

